---
title: INBO protocols
date: '`r Sys.Date()`'
---

# Welkom! / Welcome!

Voor Nederlands [klik hier](#welkom).

For English [click here](#welcome).

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  eval = TRUE,
  warning = FALSE,
  error = TRUE,
  message = FALSE
)
library(htmlwidgets)
library(slickR)
library(reactable)
```


```{r}
slickR(obj = c("voorbeeld1.png", "voorbeeld2.jpeg", "voorbeeld3.png",
               "voorbeeld4.jpg", "voorbeeld5.png", "voorbeeld6.jpg"),
       width = "95%") +
  settings(dots = TRUE, adaptiveHeight = TRUE,
           autoplay = TRUE)
```


# Nederlandstalige richtlijnen

### Welkom!

Protocollen verhogen de kwaliteit van veldwerk en labowerk en daarmee de kwaliteit van het wetenschappelijk onderzoek, dat transparant, traceerbaar en toepasbaar moet zijn.
Ze dragen ertoe bij dat veldwerk beter geharmoniseerd is en beter herhaalbaar, reproduceerbaar en vergelijkbaar wordt. Ook vanuit efficiëntie oogpunt is een standaardisatie interessant voor het Instituut voor Natuur- en Bosonderzoek (INBO).

Deze site doet dienst als een repository voor de meest recente versies van goedgekeurde, publiek raadpleegbare protocollen die op het INBO gebruikt worden.
Elk protocol kan ofwel online via deze website geraadpleegd worden (html versie) of er kan een PDF versie gedownload worden.
Aan de linkerkant vind je de navigatie naar de verschillende rubrieken.
In de NEWS sectie vind je welke (versies van) protocollen wanneer werden toegevoegd.

Deze website bevat de gepubliceerde protocollen.
Het beheer, onderhoud, goedkeuring en revisie van de bronbestanden die aan de basis liggen van deze gepubliceerde protocollen gebeurt in de [inbo/protocolsource](https://github.com/inbo/protocolsource) GitHub repository.

### Vijf soorten protocollen

We onderscheiden vijf soorten standaardprotocollen.
Het **standaard veldprotocol** (`sfp`) wordt gebruikt om de stappen uit te leggen die een veldwerker in het veld nodig heeft om een specifieke veldactiviteit uit te voeren.
Het tweede type is het **standaard instrumentprotocol** (`sip`) dat het gebruik en het onderhoud van toestellen uitlegt.
Het derde type is het **standaard analytisch protocol** (`sap`) dat in het laboratorium of in het veld wordt gebruikt om een bepaalde laboratorium- of veldanalytische procedure uit te leggen waarbij fysische of chemische metingen worden verricht.
Het vierde type staat voor **standaard operationele procedure** (`sop`) en is voor alles wat niet in een van de vorige types past.

Het vijfde type is een categorie die gereserveerd is voor **projectspecifieke protocollen** (`spp`).
Dit laatste type laat een snelle ontwikkeling en gebruik van protocollen binnen een projectcontext toe.
In het kader van een project kan het soms nodig zijn om een protocol van het type `sfp` opnieuw te gebruiken, maar met iets andere instellingen die specifiek zijn voor het project.
Dit kan worden gedaan door een gepubliceerd `sfp` protocol op te nemen als een `spp` subprotocol waarin de afwijkingen van de standaardinstellingen worden gegeven.


### Protocolidentificatie


De eerder uiteengezette typologie van protocollen wordt verder uitgewerkt in een zogenaamde protocolcode.
Een **protocolcode** bestaat uit een voorvoegsel dat verwijst naar het *protocoltype* (drie tekens), een *protocolnummer* (drie cijfers) en een *taalcode* (twee tekens), alle gescheiden door een koppelteken.

| type         | thema      | thema nummer | protocol-code |
|:-------------|:-----------|:-------------|:--------------|
| veld         | generiek   | 0            | sfp-0##-nl    |
| veld         | water      | 1            | sfp-1##-nl    |
| veld         | lucht      | 2            | sfp-2##-nl    |
| veld         | bodem      | 3            | sfp-3##-nl    |
| veld         | vegetatie  | 4            | sfp-4##-nl    |
| veld         | soorten    | 5            | sfp-5##-nl    |
| Instrument   |            |              | sip-###-nl    |
| operationeel |            |              | sop-###-nl    |
| Analyse      |            |              | sap-###-nl    |
| Project      |            |              | spp-###-nl    |

De drie cijfers van de protocolcode worden het *protocolnummer* genoemd.
Bij themaprotocollen komt het eerste cijfer van het protocolnummer overeen met het themanummer.
De `##` geven een oplopend nummer aan.
Zo zal het eerste veldprotocol voor "thema water" de protocolcode `sfp-101-en` hebben.
Het `s*p-###` deel van de protocolcode kan worden beschouwd als een code die één op één overeenkomt met de titel van het protocol (wanneer de taal buiten beschouwing wordt gelaten).

De laatste twee karakters identificeren de taal waarin het protocol is geschreven.
Dit kan het Nederlands (`nl`) of het Engels (`en`) zijn.

Elk protocol heeft dus een unieke protocolcode, maar er kunnen meerdere versies van een protocol zijn.
Om onderscheid te maken tussen verschillende versies van een protocol, hebben we een versienummer nodig.
Het **versienummer** dat we gebruiken is van de vorm `YYYY.NN`.
`YYYY` geeft het jaar aan waarin de versie van het protocol is uitgebracht.
Het `NN` is een getal dat de volgorde van uitgave binnen dat jaar aangeeft (beginnend met 01).
Het versienummer identificeert dus op unieke wijze een specifieke versie van een specifiek protocol, waarnaar verwezen kan worden om aan te geven welke versie van welk protocol gebruikt is (zonder dat de protocolcode vermeld hoeft te worden).
Hier is een fictief voorbeeld:

- versie `2022.01` (identificeert een eerste versie van `sfp-401-en`)
- versie `2022.02` (identificeert een eerste versie van `sfp-001-en`)
- versie `2022.03` (identificeert een tweede versie van `sfp-401-en`)

Op de NEWS-pagina staan de versienummers (met de recentste protocollen bovenaan) en wordt aangegeven bij welke protocollen ze horen.
De website heeft ook een overzichtspagina voor elk protocol (zoals geïdentificeerd door de protocolcode) waar alle versies van dat protocol staan.


# English guidelines

### Welcome!

Protocols increase the quality of fieldwork and labwork and thus the quality of scientific research, which must be transparent, traceable and applicable.
They help make field- and labwork more harmonized and more repeatable, reproducible and comparable.
Standardization is also interesting for the Institute for Nature and Forest
Research (INBO) from an efficiency point of view.

This site serves as a repository for the latest versions of approved, publicly available protocols used at the INBO.
Each protocol can either be accessed online through this website (html version) or a PDF version can be downloaded.
On the left side you will find the navigation to the different sections.
In the NEWS section you can find which (versions of) protocols were added when.

This website contains the published protocols.
The management, maintenance, approval and revision of the source files underlying these published protocols is done in the [inbo/protocolsource](https://github.com/inbo/protocolsource) GitHub repository.

### Five types of protocols

We distinguish five types of standard protocols.
The first type is the **standard field protocol** (`sfp`) which is used to explain the steps that are needed in the field by a field worker to execute a specific field activity.
The second type is the **standard instrument protocol** (`sip`) which explains the use and maintenance of instruments.
The third type is the **standard analytical protocol** (`sap`) which is used in the laboratory or the field to explain a particular lab or field-analytical procedure involving physical or chemical measurements.
The fourth type stands for **standard operating procedure** (`sop`) and is for everything that does not fit in one of the previous types.

The fifth type is a category reserved for **project-specific protocols** (`spp`).
The latter is more of a convenience to aid quick development and employment of protocols within a project context.
Within the scope of a project, it might sometimes be necessary to re-use a `sfp` type protocol but with slightly different settings which are specific to the project.
This can be done by including a published `sfp` protocol as an `spp` subprotocol where the deviations from the default settings are given.


### Protocol identification

The previously explained typology of protocols is further detailed in a so-called protocol-code.
A **protocol-code** consists of a prefix referring to the *protocol-type* (three characters), a *protocol-number* (three digits) and a *language tag* (two characters), all separated by a hyphen.

| type       | theme      | theme_number | protocol-code |
|:-----------|:-----------|:-------------|:--------------|
| field      | generic    | 0            | sfp-0##-en    |
| field      | water      | 1            | sfp-1##-en    |
| field      | air        | 2            | sfp-2##-en    |
| field      | soil       | 3            | sfp-3##-en    |
| field      | vegetation | 4            | sfp-4##-en    |
| field      | species    | 5            | sfp-5##-en    |
| instrument |            |              | sip-###-en    |
| operating  |            |              | sop-###-en    |
| analysis   |            |              | sap-###-en    |
| project    |            |              | spp-###-en    |

The three digits of the protocol-code will be referred to as the *protocol-number*.
In case of thematic protocols, the first digit of the protocol-number corresponds with the theme-number.
The `##` indicates an incremental number.
For instance, the first field protocol for "theme water" will have protocol code `sfp-101-en`.
The `s*p-###` part of the protocol code can be thought of as a code that corresponds one-to-one with the title of the protocol (when ignoring language).

The final two characters identify the language the protocol is written in.
This can be either Dutch (`nl`) or English (`en`).

Each protocol thus has a unique protocol-code, but there can be multiple versions of a protocol.
To discriminate between different versions of protocol, we need a version number.
The **version number** that we use is of the form `YYYY.NN`.
`YYYY` indicates the year in which the version of the protocol was released.
The `NN` is a number that indicates the order of release within that year (starting with 01).
The version number will therefore uniquely identify a specific version of a specific protocol, which can be refered to without the need to refer to the protocol-code.
Here's a fictive example:

- version `2022.01` (identifies a first version of `sfp-401-en`)
- version `2022.02` (identifies a first version of `sfp-001-en`)
- version `2022.03` (identifies a second version of `sfp-401-en`)

The NEWS page lists the version numbers (in descending chronological order) and identifies which protocols they belong to.
The website also has an overview page for each protocol (as identified by protocol-code) where all versions of that protocol are listed.

# Standard field protocols (sfp)

```{r}
inbo_hoofd <- rgb(192, 67, 132, maxColorValue = 255)
inbo_achtergrond <- rgb(243, 243, 243, maxColorValue = 255)

table_theme <- reactableTheme(
    borderColor = inbo_hoofd)

row_style <- JS("
  function(rowInfo, state) {
  if (rowInfo.level > 0) {
  if (rowInfo.values['striping']) {
      return { background: 'rgba(192, 67, 132, 0.1)' }
    }
  }
  }
")
```


```{r sfp}
sfp <- read.csv("sfp.csv")
reactable(sfp,
          columns = list(
            version = colDef(
              html = TRUE,
              filterable = TRUE,
              rowHeader = TRUE,
              cell = function(value, index) {
                sprintf('<a href="%s" target="_blank">%s</a>',
                        sfp$address[index], value)
              }
            ),
            code = colDef(filterable = TRUE),
            title = colDef(filterable = TRUE),
            address = colDef(show = FALSE),
            striping = colDef(show = FALSE)
            ),
          defaultSorted = list(theme = "asc", code = "asc", version = "desc"),
          sortable = FALSE,
          defaultPageSize = 10,
          groupBy = "theme",
          paginateSubRows = TRUE,
          defaultExpanded = TRUE,
          theme = table_theme,
          rowStyle = row_style,
          borderless = TRUE
          )
```


# Standard instrument protocols (sip)

```{r sip}
sip <- read.csv("sip.csv")

reactable(sip,
          columns = list(
            version = colDef(
              html = TRUE,
              filterable = TRUE,
              rowHeader = TRUE,
              cell = function(value, index) {
                sprintf('<a href="%s" target="_blank">%s</a>',
                        sip$address[index], value)
              }
            ),
            code = colDef(filterable = TRUE),
            title = colDef(filterable = TRUE),
            address = colDef(show = FALSE),
            theme = colDef(show = FALSE),
            striping = colDef(show = FALSE)
            ),
          defaultSorted = list(theme = "asc", code = "asc", version = "desc"),
          sortable = FALSE,
          defaultPageSize = 10,
          paginateSubRows = TRUE,
          defaultExpanded = TRUE,
          theme = table_theme,
          rowStyle = row_style,
          borderless = TRUE
          )
```

# Standard analytical protocols (sap)

```{r sap}
sap <- read.csv("sap.csv")

reactable(sap,
          columns = list(
            version = colDef(
              html = TRUE,
              filterable = TRUE,
              rowHeader = TRUE,
              cell = function(value, index) {
                sprintf('<a href="%s" target="_blank">%s</a>',
                        sap$address[index], value)
              }
            ),
            code = colDef(filterable = TRUE),
            title = colDef(filterable = TRUE),
            address = colDef(show = FALSE),
            theme = colDef(show = FALSE),
            striping = colDef(show = FALSE)
            ),
          defaultSorted = list(theme = "asc", code = "asc", version = "desc"),
          sortable = FALSE,
          defaultPageSize = 10,
          paginateSubRows = TRUE,
          defaultExpanded = TRUE,
          theme = table_theme,
          rowStyle = row_style,
          borderless = TRUE
)
```

# Standard operating procedures (sop)

```{r sop}
sop <- read.csv("sop.csv")
reactable(sop,
          columns = list(
            version = colDef(
              html = TRUE,
              filterable = TRUE,
              rowHeader = TRUE,
              cell = function(value, index) {
                sprintf('<a href="%s" target="_blank">%s</a>',
                        sop$address[index], value)
              }
            ),
            code = colDef(filterable = TRUE),
            title = colDef(filterable = TRUE),
            address = colDef(show = FALSE),
            theme = colDef(show = FALSE),
            striping = colDef(show = FALSE)
            ),
          defaultSorted = list(theme = "asc", code = "asc", version = "desc"),
          sortable = FALSE,
          defaultPageSize = 10,
          paginateSubRows = TRUE,
          defaultExpanded = TRUE,
          theme = table_theme,
          rowStyle = row_style,
          borderless = TRUE
)
```

# Project-specific protocols (spp)

```{r spp}
spp <- read.csv("spp.csv")

reactable(spp,
          columns = list(
            version = colDef(
              html = TRUE,
              filterable = TRUE,
              rowHeader = TRUE,
              cell = function(value, index) {
                sprintf('<a href="%s" target="_blank">%s</a>',
                        spp$address[index], value)
              }
            ),
            code = colDef(filterable = TRUE),
            title = colDef(filterable = TRUE),
            address = colDef(show = FALSE),
            striping = colDef(show = FALSE),
            theme = colDef(name = "project")
            ),
          defaultSorted = list(theme = "asc", code = "asc", version = "desc"),
          sortable = FALSE,
          groupBy = "theme",
          defaultPageSize = 10,
          paginateSubRows = TRUE,
          defaultExpanded = TRUE,
          theme = table_theme,
          rowStyle = row_style,
          borderless = TRUE
)
```
